Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
intro shigeki edits

A single genome holds the instructions to generate the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Although many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by mutations in chromatin regulators[@gabrieleChromatinBasisNeurodevelopmental2018]. Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017], such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. Very few studies have investigated the misexpression of tissue-specific genes in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if this partial loss of brain identity contributes to neurodevelopmental impairments.

To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the types of genes dysregulated and the molecular mechanisms governing their de-repression. In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C). KDM5C, , also known as SMCX or JARID1C, is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose predominant features include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017]. It is currently unknown if this dysregulation of brain tissue identity further impairs _Kdm5c_-KO neurodevelopment and if ectopic gene expression within the _Kdm5c_-KO brain is unique to testis genes.

- one key process chrom reg are involved in are is distinguishing bw germ + soma cell identity

Distinguishing between the germline (cells that pass on their genetic material, e.g. sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis. In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary. Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are typically uniquely expressed in germ cells. While the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_. KDM5C was highly enriched at germline gene promoters in EpiLCs but only bound to a subset of germline genes expressed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA methylation in EpiLCs through H3K4me2/3 removal. Thus, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@

- **note: should I compare amygdala and hippocampus? scandaglia only looked at hippocampus**

>>>>

A single genome holds the instructions to generate the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. While many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by mutations in chromatin regulators[@gabrieleChromatinBasisNeurodevelopmental2018]. Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if partial loss of brain identity contributes to neurodevelopmental impairments.

To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the types of genes dysregulated and the molecular mechanisms governing their de-repression. In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C). KDM5C, , also known as SMCX or JARID1C, is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose predominant features include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017]. It is currently unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development this begins, and if other tissue-specific genes are also aberrantly expressed with KDM5C loss.

The testis contains both germ cells (cells that pass on their genetic material, e.g. sperm) and somatic cells (cells that perform all other bodily functions). Distinguishing between the germline and the soma is a key feature of multicellularity and that typically occurs during early embryogenesis. In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency. Initially, repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], are placed at germline gene promoters in embryonic stem cells and are then decorated with DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] in the post-implantation embryo. How KDM5C interacts with known germline gene silencers during this time point is currently unknown. Additionally, most studies have focused on select genes important for early germ cell development rather than germline-enriched genes as a whole given the lack of a curated list of germline-enriched genes. Thus, it is unclear if the mechanism of repression differs for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes. Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary. Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes. We demonstrated testis and ovary-enriched genes genes are germline genes by generating a list of germline enriched-genes using germ cell-depleted RNA sequencing datasets. Using this curated list, we found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis. Additionally, KDM5C was highly enriched at germline gene promoters in EpiLCs but only bound to a subset of germline genes expressed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs. Thus, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@
